---
title: "Markdown_Report"
author: "Lauren"
date: "April 4, 2020"
output: word_document
params:
  file_path: "./Output/wildfire_2016.csv"
  year: 2016

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r load data}
library(dplyr)
library(ggplot2)
library(lubridate)
library(knitr)
#library(tidyverse)

wildfire_table<-readr::read_csv(params$file_path, col_types=list("c","n","c","c","c","n","n","n","c","c","n","n","n","c","c","n"))
```


The main fields we will be focusing on are `FIRE_TYPE`,`FIRE_CAUSE`,`FIRE_YEAR `,`CURRENT_SIZE`, and `IGNITION_DATE`

FIRE_CAUSE is the suspected cause of the fire, e.g., Lightning, Person.

FIRE_TYPE defines the type of incident that occurred, e.g., Fire of Note, New, Modified/Monitored, Full Response.

FIRE_YEAR represents the fiscal year, April 1 to March 31.

IGNITION_DATE is the discovery date of the fire.

CURRENT_SIZE is the estimated size of the area within the outside perimeter of the incident, in hectares.


##Exploratory Analysis of Wildfires

Let's set out to investigate these questions first:

1) How many entries of fires do we have for a given year?

2) What is the average,minimum and maximum `SIZE_HA` that we have for our data?

3) What is the count of fires in a year?



```{r exploratory analysis}
wild_sum<-wildfire_table%>%summarize(average=mean(SIZE_HA,na.rm=T),min=min(SIZE_HA,na.rm=T),max=max(SIZE_HA,na.rm=T))

knitr::kable(wild_sum, format.args=list(big.mark=","))
```

There were `r nrow(wildfire_table)` fires over the time period from `r min(wildfire_table$IGN_DATE)` and `r max(wildfire_table$IGN_DATE)` 

```{r bar month}
wildfire_table%>%
  mutate(IGN_MONTH=month(IGN_DATE,label=TRUE,abbr = FALSE))%>%
  group_by(IGN_MONTH)%>%
  count()%>%
  ggplot(aes(x=IGN_MONTH,y=n,group=1))+
  geom_col(alpha=0.8,fill="darkgreen")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


```{r barchart_theme}
ggplot(data=wildfire_table, aes(x=FIRE_TYPE))+
  scale_fill_brewer(type="qual",palette = "Dark2")+
  geom_bar(aes(fill=FIRE_CAUSE))+
  labs(x="Fire Type", y="Total Entries",fill="Fire Cause",title = "Fires Across British Columbia")+
  coord_flip()+
  theme(legend.position = "bottom")
  


```


