---
title: "Session4_Exercise"
author: "Lauren"
date: "April 8, 2020"
output: 
  html_document:
    theme: cosmo
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE,message = FALSE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(plotly)
library(leaflet)
#install.packages("DT")
library(DT)
library(gganimate)
```


```{r load data}

wildfire_table<-readr::read_csv("./Output/wildfire_clean.csv", col_types=list("c","n","c","c","c","n","n","n","c","c","n","n","n","c","c","n"))
```


We're taking our chart from last session and making it interactive. We can do this simply by using `ggplotly` and our plot.

# Fires in British Columbia {.tabset}

## First plotly

```{r barchart_theme}
cause_plot=ggplot(data = wildfire_table, aes(x = FIRE_TYPE)) +
  geom_bar(aes(fill = FIRE_CAUSE)) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  labs(x = "Fire Cause",
  y = "Total Entries",
  fill = "Fire Cause",
  title = "Fires Across British Columbia") +
  scale_y_continuous(labels = scales::comma) +
  theme(
  legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.margin = margin(t = 15),
  axis.title.y = element_text(margin = margin(r = 20)),
  axis.title.x = element_text(margin = margin(t = 20))
  )
  


fig<-ggplotly(cause_plot)
fig
```

## Tool tips

Specify text in the tool tip:
`\n` is a hack to get a new line in the tooltip

```{r tooltip}
cause_plot=ggplot(data = wildfire_table, aes(x = FIRE_TYPE, text=
                                               paste0("Fire Type:", FIRE_TYPE, "\n","Fire Cause:",FIRE_CAUSE,"\n"))) +
  geom_bar(aes(fill = FIRE_CAUSE)) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  labs(x = "Fire Cause",
  y = "Total Entries",
  fill = "Fire Cause",
  title = "Fires Across British Columbia") +
  scale_y_continuous(labels = scales::comma) +
  theme(
  legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.margin = margin(t = 15),
  axis.title.y = element_text(margin = margin(r = 20)),
  axis.title.x = element_text(margin = margin(t = 20))
  )
  


fig<-ggplotly(cause_plot, tooltip = "text")
fig

```

Specify the interaction and what happens with `dragmode = "pan"`

# Wildfires in 2018 {.tabset}

## Wildfires by Month

```{r wildfires2018}
wild_2018<-wildfire_table%>%filter(FIRE_YEAR==2018)

wild_month=wild_2018%>%
  mutate(IGN_MONTH=month(IGN_DATE,label=TRUE,abbr = FALSE))%>%
  group_by(IGN_MONTH)%>%
  count()%>%
  ggplot(aes(x=IGN_MONTH,y=n,group=1))+
  geom_point()+
  geom_line()+
  labs(x="Month",y="Number of Wildfires")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

fig=ggplotly(wild_month)%>% layout(dragmode = "pan")

fig

```

```{r data table}

wild_2018_sub<-wild_2018%>%select(IGN_DATE,FIRE_CAUSE,FIRE_TYPE,SIZE_HA)
datatable(wild_2018_sub, options = list(pageLength = 5), colnames = c("Ignition Date", "Fire Cause","Fire Type","Size (ha)"),filter="top")

```

## Daily Wild Fires 

```{r lines}
daily =  wildfire_table%>%
  filter(between(year(IGN_DATE), 2017,2018))%>%
  group_by(IGN_DATE=as_date(IGN_DATE))%>%
  count()%>%
  ggplot(aes(x=IGN_DATE,y=n,group=1))+
  geom_point()+
  geom_line()+
  labs(x="Date",y="Number of Wildfires")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggplotly(daily, dynamicTicks = TRUE) %>%
  rangeslider() %>%
  layout(hovermode = "x")

```

Animation

```{r monthly_data}

monthly_data = 
  wildfire_table%>%
  filter(between(year(IGN_DATE), 2017,2018))%>%
  mutate(IGN_YEAR=year(IGN_DATE),IGN_MONTH=month(IGN_DATE,label=TRUE,abbr = FALSE))%>%
  group_by(IGN_YEAR,IGN_MONTH)%>%
  count()
```

```{r,gganimate}
monthly_plot = 
  ggplot(data=monthly_data,aes(x=IGN_MONTH,y=n,group=1))+
  geom_col(fill="#54278f",alpha=0.4)+
  labs(x="Date",y="Number of Wildfires")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

monthly_plot +
  ggtitle("Year: {frame_time}")+
   transition_time(as.integer(IGN_YEAR))+
  ease_aes('linear')
  

```

```{r lines2}
rshp_monthly<-monthly_data %>%
  pivot_wider(names_from=IGN_YEAR, values_from = n,names_prefix="Y_")%>%
  ungroup()
rshp_monthly$Y_2018[2]=0
plotly_figure<-plot_ly(data = rshp_monthly, x = ~IGN_MONTH)%>%
  add_trace(y=~Y_2017,name = '2017',type='scatter',mode='lines+markers')%>%
  add_trace(y=~Y_2018,name = '2018',type='scatter', mode = 'lines+markers')
plotly_figure

```


## Facets

```{r facets}

wildfire_table<-wildfire_table%>%
  mutate(IGN_MONTH=month(IGN_DATE,label=TRUE,abbr = FALSE))

wildfire_year<-wildfire_table%>%
  group_by(IGN_YEAR=year(IGN_DATE),IGN_MONTH)%>%
  summarize(n=n())%>%
  filter(between(IGN_YEAR,2008,2017))

facet_plot<-wildfire_year%>%
  ggplot(aes(x=IGN_MONTH,y=n,group=1))+
  geom_point()+
  geom_line()+
  labs(x="Month",y="Number of Wildfires")+
  facet_wrap(nrow=5,vars(IGN_YEAR))+
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

ggplotly(facet_plot)

```



# Leaflet

```{r spatial data}
library(sf)

wild_shp<-st_as_sf(wildfire_table,coords=c("LONGITUDE","LATITUDE"),crs=4326)
wild_shp<-wild_shp %>%filter(year(IGN_DATE)==2017)

```

```{r leaflet}
library(leaflet)
leaflet(wild_shp) %>%
  #setView(174.764, -36.877, zoom = 16) %>%
  addTiles()%>%
  addMarkers(clusterOptions = markerClusterOptions())


pal <- colorFactor(c("navy", "red","grey"), domain = c("Person", "Lightning","Unknown"))

leaflet(wild_shp) %>% 
  addTiles() %>%
  addProviderTiles(providers$Stamen.Toner, group = "Toner") %>%
  addProviderTiles(providers$Stamen.TonerLite, group = "Toner Lite")%>%
   addLayersControl(
    baseGroups = c("OSM", "Toner", "Toner Lite"))%>%
  addCircleMarkers(
    radius = ~sqrt((SIZE_HA/10000)/pi),
    color = ~pal(FIRE_CAUSE),
    stroke = FALSE, fillOpacity = 0.5,
    popup= ~paste(sep = "<br/>",FIRE_CAUSE, "Size:",SIZE_HA," ha")
  )
```



