---
title: "Session3_exercise"
author: "Lauren"
date: "April 4, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


```{r load data}
library(dplyr)
library(ggplot2)
library(lubridate)
#library(tidyverse)

wildfire_table<-readxl::read_excel("./Data/wildfires_table.xlsx", col_types=c("text","numeric","text","text","text","numeric","numeric","numeric","text","text","numeric","numeric","numeric","text","text","numeric"))
```



```{r reformat date}
wildfire_table<-wildfire_table %>%
    mutate(IGN_DATE = parse_date_time(IGN_DATE,"ymd"))

```

Uh-oh, our other field `FIREYEAR` has a minimum of 1950. What about dates that didn't match our format? We can investigate this further by:

```{r fix date}

wildfire_table %>% filter(year(IGN_DATE)==1930)

wildfire_table %>%filter(is.na(IGN_DATE))

wildfire_table<-wildfire_table%>%filter(!year(IGN_DATE)==1930)

```


The *duplicate* in FIRE_TYPE should probably be removed:

```{r remove dup}
wildfire_table<-wildfire_table%>%filter(!FIRE_TYPE=="Duplicate")


```



#Output Data

In the chunk below we are going to export two new wildfire data sets for different years to use in the Markdown_Report.Rmd. The first subset of data will be from the year 2016 and the second from 2017. We are going to output these into the `Output` folder in your working directory/project folder. Create this folder if it doesn't exist.

Additionally, we are going to output ALL the data into a new csv as "wildfire_clean.csv" so we are dealing with only the data that we have cleaned from above.

```{r outputclean}

write_csv(wildfire_table,"./Output/wildfire_clean.csv")

```

```{r output1}

output1<- wildfire_table %>%filter(year(IGN_DATE)==2016)

write_csv(output1,"./Output/wildfire_2016.csv")


```


Using the example above to guide you, generate the outputs for 2017.

```{r output2}
write_csv()

```



